Carrier phase and amplitude estimation for phase shift keying using pilots and data (Appendix)

### 2012-08-20 paper1.tex
Added some edits

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly \emph{symbol error rate}, or \emph{bit error rate}.  Whilst estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite approach here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated recieved symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent reciever operations, such as decoding.  Consequently, our metric of performance will not be symbol or bit error rate, it will be mean square error (MSE) between complex amplitude $a_0$ and its estimator $\hat{a}$, that is, our metric is $\expect\abs{a_0 - \hat{a}}^2$ where $\expect$ denotes the expected value. It will actually be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $\expect(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\expect\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $\expect(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorms that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle^2$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle^2$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho_0}$ is not a consistent estimator of the amplitude $\rho_0$.  The bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  Due to space constraints, the lemmas needed by these proofs have been placed in the supplementary paper~\cite{McKilliam_leastsqPSKpilotsdata_2012appendix}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic variance can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{W}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{W}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
\[
f_{X,Y}(x,y) = \frac{f_Z(z)}{2\pi z} \qquad \text{where $z = \sqrt{x^2 + y^2}$.}
\]
Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
\begin{align*}
 \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
 &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
 &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
 \end{align*}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $\abs{G(x)}$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  This assumption asserts that $\abs{G(x)} \leq \abs{G(0)}$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence such that $\abs{G(x_i)} \rightarrow \sabs{G(0)}$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  Although we will not prove it here, this assumption is not only sufficient, but also nescessary, for if $\abs{G(x)}$ is uniquely maximised at some point $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $\abs{G(x)}$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  The next Lemma describes a class of circularly symmetric distributions for which $\abs{G(x)}$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is apart of this class.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable with pdf $f_{X,Y}(x,y)$ of the real and imaginary parts nonincreasing with $x^2 + y^2$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $\abs{h_1(x)}$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $\abs{h_2(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

>>>>

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly \emph{symbol error rate}, or \emph{bit error rate}.  Whilst estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite approach here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated recieved symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent reciever operations, such as decoding.  Consequently, our metric of performance will not be symbol or bit error rate, it will be square error between complex amplitude $a_0$ and its estimator $\hat{a}$, that is, our metric is $\abs{a_0 - \hat{a}}^2$. It will actually be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorms that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle^2$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle^2$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  Due to space constraints, the lemmas needed by these proofs have been placed in the supplementary paper~\cite{McKilliam_leastsqPSKpilotsdata_2012appendix}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{W}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{W}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
\begin{align*}
 \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
 &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
 &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
 \end{align*}
If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
\[
f_{X,Y}(x,y) = \frac{f_Z(\sqrt{x^2 + y^2})}{2\pi \sqrt{x^2 + y^2}}.
\]

@@

A key assumption in Theorem~\ref{thm:consistency} is that $\abs{G(x)}$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  This assumption asserts that $\abs{G(x)} \leq \abs{G(0)}$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $\abs{G(x_i)} \rightarrow \sabs{G(0)}$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  Although we will not prove it here, this assumption is not only sufficient, but also nescessary, for if $\abs{G(x)}$ is uniquely maximised at some point $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $\abs{G(x)}$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  The next Lemma describes a class of circularly symmetric distributions for which $\abs{G(x)}$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is apart of this class.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $\abs{h_1(x)}$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $\abs{h_2(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

### 2012-08-20 paper2.tex
started filling in the simulation section.
Added some edits

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data (Appendix)}
\author{Robby McKilliam, Andre Pollok, Bill Cowley, I. Vaughan L. Clarkson and Barry Quinn
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.
}}

@@

\begin{lemma}\label{lem:convtoexpGlamL} 
$\sabs{G(\hat{\lambda}_L)} \rightarrow \sabs{G(0)}$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Since $\sabs{G(x)}$ is uniquely maximised at $x=0$,
\[
0 \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)},
\]
and since $\hat{\lambda}_L$ is the maximiser of $\abs{G_L(x)}$,
\[ 
0 \leq \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}.
\]
Thus,
\begin{align*}
0 &\leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} \\ 
& \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} + \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}\\
& \leq |G(0) - G_L(0)| + |G_L(\hat{\lambda}_L) - G(\hat{\lambda}_L)| \\
&\leq 2\sup_{\lambda \in [-\pi, \pi)} \sabs{G_L(\lambda) - G(\lambda)},
\end{align*}
and the last line converges amost surely to zero by Lemma~\ref{lem:uniflawGGL}.
\end{IEEEproof}

@@

\begin{lemma}\label{lem:supsupDLlambda} For any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{2\pi}{N}\expect R_i \right\} = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that
\begin{align*}
&\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&=  \abs{G_L(\lambda) - T_L(\lambda) - G_L(\lambda_n) + T_L(\lambda_n)} \\
&\leq  \abs{G_L(\lambda) - G_L(\lambda_n)} + \abs{\expect G_L(\lambda) - \expect G_L(\lambda_n)} \\
&\leq \abs{G_L(\lambda) - G_L(\lambda_n) } + \expect \abs{G_L(\lambda) - G_L(\lambda_n)},
\end{align*}
the last line following from Jensens inequality.  Put
\[
C_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \abs{G_L(\lambda) - G_L(\lambda_n) },
\]
so that
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)} \leq C_L + \expect C_L,
\]
because 
\[
 \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \expect \abs{G_L(\lambda) - G_L(\lambda_n)} \leq \expect C_L
\]
follows by moving expectation outside the supremum.  Lemma~\ref{lem:CL} shows that $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$ and also that
\[
\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\]
So,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} (C_L + \expect C_L) > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

\begin{lemma}\label{lem:CL} The following statements hold:
\begin{enumerate}
\item $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$,
\item for any $\epsilon > 0$, $\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0$.
\end{enumerate}
\end{lemma}
\begin{IEEEproof}
If $\lambda \in L_n$, then $\lambda = \lambda_n + \delta$ with $\delta < \tfrac{2\pi}{N}$, and from Lemma~\ref{lem:coslipshitz},
\begin{align*}
&\abs{\cos(\lambda + \Phi_i) - \cos(\lambda_n + \Phi_i)} \leq \tfrac{2\pi}{N}, \;\;\; \text{and} \\
&\abs{\cos\sfracpart{\lambda + \Phi_i} - \cos\sfracpart{\lambda_n + \Phi_i}} \leq \tfrac{2\pi}{N}.
\end{align*}
Because this results does not depend on $n$,
\[
\sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \leq R_i \frac{2\pi}{N}
\]
for all $i = P \cup D$.  Now,
\begin{align*}
C_L &= \sup_{n=1,\dots,N}\sup_{\lambda \in L_n}\abs{\frac{1}{L}\sum_{i \in P \cup D} Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{1}{L}\sum_{i \in P \cup D} \sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i .
\end{align*}
Thus, 
\[
\expect C_L \leq \expect \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i = \tfrac{2\pi}{N}\expect R_1
\]
and the first statement holds.  Now, 
\[
\frac{2\pi}{N L}\sum_{i \in P \cup D} R_i \rightarrow \frac{2\pi}{N}\expect R_1
\] 
almost surely to $\frac{2\pi}{N}\expect R_1$ as $L \rightarrow\infty$ by the strong law of large numbers, and so, for any $\epsilon > 0$,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

>>>>

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data (Appendix)}
\author{Robby McKilliam, Andre Pollok, Bill Cowley, I. Vaughan L. Clarkson and Barry Quinn
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.
}}

@@

\begin{lemma}\label{lem:convtoexpGlamL} 
$\sabs{G(\hat{\lambda}_L)} \rightarrow \sabs{G(0)}$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Since $\sabs{G(x)}$ is uniquely maximised at $x=0$,
\[
0 \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)},
\]
and since $\hat{\lambda}_L$ is the maximiser of $\abs{G_L(x)}$,
\[ 
0 \leq \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}.
\]
Thus,
\begin{align*}
0 &\leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} \\ 
& \leq \sabs{G(0)} - \sabs{G(\hat{\lambda}_L)} + \sabs{G_L(\hat{\lambda}_L)} - \sabs{G_L(0)}\\
& \leq |G(0) - G_L(0)| + |G_L(\hat{\lambda}_L) - G(\hat{\lambda}_L)| \\
&\leq 2\sup_{\lambda \in [-\pi, \pi)} \sabs{G_L(\lambda) - G(\lambda)},
\end{align*}
and the last line converges almost surely to zero by Lemma~\ref{lem:uniflawGGL}.
\end{IEEEproof}

@@

\begin{lemma}\label{lem:supsupDLlambda} For any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_i \right\} = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that
\begin{align*}
&\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&=  \abs{G_L(\lambda) - T_L(\lambda) - G_L(\lambda_n) + T_L(\lambda_n)} \\
&\leq  \abs{G_L(\lambda) - G_L(\lambda_n)} + \abs{\expect G_L(\lambda) - \expect G_L(\lambda_n)} \\
&\leq \abs{G_L(\lambda) - G_L(\lambda_n) } + \expect \abs{G_L(\lambda) - G_L(\lambda_n)},
\end{align*}
the last line following from Jensens inequality.  Put
\[
C_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \abs{G_L(\lambda) - G_L(\lambda_n) },
\]
so that
\begin{align*}
V_L &= \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&\leq  C_L +  \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \expect\abs{G_L(\lambda) - G_L(\lambda_n) } \\
&\leq C_L + \expect C_L,
\end{align*}
where the last line follows because $\sup \expect\vert \dots\vert \leq \expect \sup \vert \dots \vert$.  Lemma~\ref{lem:CL} shows that $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$ and also that
\[
\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\]
Thus,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} (C_L + \expect C_L) > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

\begin{lemma}\label{lem:CL} The following statements hold:
\begin{enumerate}
\item $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$,
\item for any $\epsilon > 0$, $\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0$.
\end{enumerate}
\end{lemma}
\begin{IEEEproof}
If $\lambda \in L_n$, then $\lambda = \lambda_n + \delta$ with $\delta < \tfrac{2\pi}{N}$, and from Lemma~\ref{lem:coslipshitz},
\begin{align*}
&\abs{\cos(\lambda + \Phi_i) - \cos(\lambda_n + \Phi_i)} \leq \tfrac{2\pi}{N}, \;\;\; \text{and} \\
&\abs{\cos\sfracpart{\lambda + \Phi_i} - \cos\sfracpart{\lambda_n + \Phi_i}} \leq \tfrac{2\pi}{N}.
\end{align*}
Because this results does not depend on $n$,
\[
\sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \leq R_i \frac{2\pi}{N}
\]
for all $i = P \cup D$.  Now,
\begin{align*}
C_L &= \sup_{n=1,\dots,N}\sup_{\lambda \in L_n}\abs{\frac{1}{L}\sum_{i \in P \cup D} Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{1}{L}\sum_{i \in P \cup D} \sup_{n=1,\dots,N}\sup_{\lambda \in L_n} \abs{Z_i(\lambda) - Z_i(\lambda_n)} \\
&\leq \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i .
\end{align*}
Thus, 
\[
\expect C_L \leq \expect \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i = \frac{2\pi}{N}\expect R_1
\]
and the first statement holds.  Now, 
\[
\frac{2\pi}{N L}\sum_{i \in P \cup D} R_i \rightarrow \frac{2\pi}{N}\expect R_1
\] 
almost surely to $\frac{2\pi}{N}\expect R_1$ as $L \rightarrow\infty$ by the strong law of large numbers, and so, for any $\epsilon > 0$,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} \frac{2\pi}{N L}\sum_{i \in P \cup D} R_i > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}
